Designing and Modeling a Browser-Based Distributed Evolutionary Computation System

### 2015-05-09 GECCO/wk0705.tex
Version ready for uploading

Most theory and experimentation in parallel computing has been devoted to predict and
optimize the performance in systems where the number of nodes, their
connections, and the time every one is dedicating to the computation is
known in advance as it often did in the Big Science era, where labs
could easily buy 16-node clusters or subscribe to the resources of a
big grid. Even today, you can rent instances of virtual machines in
the cloud relatively cheaply and use them for evolutionary computation
experiments \cite{sherry2012flex}. Nothing beats free, which is one of
the reasons why the era of
Citizen Science already started a few years ago (with SETI@home \cite{david-seti:home} and then BOINC \cite{boinc_grid04}) and it
offers a vast amount of computational resources to be used, if only
you know how to attract them. But there is a challenge: knowing, or at least having a
ballpark, of how your algorithm is going to perform in this uncertain
environment, where none of the factors is known: neither the number of
nodes, through how they are connected, to how long are they going to
be focused on doing what you want them to. That is why some effort is
needed to first understand the dynamics of the decision to participate
in an experiment that requires you to click on a link and then stay
for a while looking at the screen (or just leave it there
running). Volunteer computing experiments were set up for the first
time in the 90s, in the shape of the SETI@home screensaver
\cite{david-seti:home} that
analyzed signals looking for patterns indicating the existence of
extraterrestrial intelligence due to budget cuts by the US
congress. They proved \cite{anderson2006computational}
that 
\begin{quote}
... [V]olunteer computing can support
applications that are significantly more data-intensive,
or have high memory and storage requirements, than
those in current projects.
\end{quote}
Evolutionary algorithms are that kind of applications. In general,
real world problems and some combinatorial optimization problems such
as the MasterMind puzzle \cite{verbatim} do require an amount of
computational resources unavailable to the average researcher. On the
other hand, one clear advantage of volunteer computing frameworks is
its availability: in principle, any one can request volunteers. How
many she is able to gather, on the other hand, is an issue that we
will try to approach in this paper.

Besides, since Amazon started selling EC2 several years ago, reliable
and scalable computing resources are also available for a low price
and on demand. Recently, Google has also refurbished its offering
lowering their prices. This means that the conjunction of free or
low-cost cloud computing engines, volunteer computing systems and
untapped capability of other people's desktop systems can be used for creating
massive, or at least potentially massive, distributed computing
experiments. These experiments can be easily set up using open-source
repository sites like GitHub\footnote{\url{http://www.github.com}} and deployed automatically to Platform as
a Service (PaaS) sites such as
Heroku\footnote{\url{https://www.heroku.com/}} or
OpenShift\footnote{\url{https://www.openshift.com/}} or even
Infrastructure as a Service (IaaS) sites such as Microsoft's Azure or
Amazon's EC2.

@@

These are best practices that have been followed in the experiments in
this paper, that first presents a new version of a platform for
distributed evolutionary algorithms (EAs) using the browser and a free (as
in free beer) backend, and second, shows the result from a
statistical point of view in order to put the basis for a model of the
meta-computer obtained by joining all volunteers and the free backend
used for the experiment. This has been one of the criterion when
choosing the different cloud services (hosting and logging, mainly):
the fact that they are free, and sustainably so
\cite{Merelo2014,jj:idc:lowcost}.

@@

\section{Results}
\label{sec:res}
%
\begin{figure*}[h!tb]
        \centering
        \includegraphics[width=0.95\linewidth]{img/times-base.png}
        \caption{Histogram of time needed to find the solution in the
          command-line client case, with (Y) or without restart (N)
          when the solution is found.}\label{fig:base}
\end{figure*}
%
In order to establish a baseline, we wrote an evolutionary algorithm
in node using the NodEO framework \cite{DBLP:conf/gecco/GuervosVGES14}
in the command line. Conditions, and, indeed, the code, were the same except for the
population, which was set to 512 individuals, and the termination
condition, which included finishing if a sufficient amount of
generations had passed without finding the solution. We did 50 experiments restarting the server,
which contains the pool of best individuals every 100 generations,
when the client had finished without finding the solution, or not;
this second case is more similar to the asynchronous nature of the
browser-based experiment. In fact, no difference was found (using
Wilcoxon test) and the average time to solution was 34.21 seconds.

@@

At the time of this writing, 23 runs were done in OpenShift and one in  
% para trabajos futuros, estas cifras de cuántas ejecuciones se han completado 
% quizás habría que presentarlas de forma diferente. Quizás indicando el tiempo
% total de "puesta online" del sistema, cómo se publicitó para atraer gente,
% y entonces ahí contar las estadísticas de cuánta gente partició, cuántas
% ejecuciones se terminaron, etc.  [Pedro]
% se terminaron todas. Lo de cuanta gente por ejecución (cuantas IPs,
% en realidad) y cuanto tardaron no lo he llegado a analizar, aunque
% tengo los datos. Se deja para trabajo futuro, puedes añadirlo a un
% issue - JJ
Heroku. The Openshift URL was mentioned more frequently than the other
so that accounts for the popularity, not any particular problem with
the algorithm in one or the other. That is why the data that we have
obtained and used below, and which is, as usual, available with a free
license, is mainly processed from the log files stored in Loggly and
in OpenShift. These 23 experiments took each one a number of hours, so
that in many cases the nodes were not accessing at the same time, but
sequentially, to the server. This means that the island model is, in
fact, behaving partly as a pool-based evolutionary algorithm
\cite{sofea:evopar2012} with some results, the best from each
generation, available to future runs. This is clearly noticed when
visualizing the algorithm: its behavior changes when it gets its first
chromosome from the pool, in many cases improving a lot.

>>>>

Most theory and experimentation in parallel computing has been devoted to predict and
optimize the performance in systems where the number of nodes, their
connections, and the time every one is dedicating to the computation is
known in advance as it often did in the Big Science era, where labs
could easily buy 16-node clusters or subscribe to the resources of a
big grid. Even today, you can rent instances of virtual machines in
the cloud relatively cheaply and use them for evolutionary computation
experiments \cite{sherry2012flex}. Nothing beats free, which is one of
the reasons why the era of
Citizen Science already started a few years ago (with SETI@home \cite{david-seti:home} and then BOINC \cite{boinc_grid04}) and it
offers a vast amount of computational resources to be used, if only
you know how to attract them. But there is a challenge: knowing, or at least having a
ballpark, of how your algorithm is going to perform in this uncertain
environment, where none of the factors is known: neither the number of
nodes, through how they are connected, to how long are they going to
be focused on doing what you want them to. That is why some effort is
needed to first understand the dynamics of the decision to participate
in an experiment that requires you to click on a link and then stay
for a while looking at the screen (or just leave it there
running). Volunteer computing experiments were set up for the first
time in the 90s, in the shape of the SETI@home screensaver
\cite{david-seti:home} that
analyzed signals looking for patterns indicating the existence of
extraterrestrial intelligence due to budget cuts by the US
congress. They proved \cite{anderson2006computational}
that 
\begin{quote}
... [V]olunteer computing can support
applications that are significantly more data-intensive,
or have high memory and storage requirements, than
those in current projects.
\end{quote}
Evolutionary algorithms are that kind of applications. In general,
real world problems and some combinatorial optimization problems such
as the MasterMind puzzle \cite{DBLP:journals/evi/Maestro-MontojoSG14} do require an amount of
computational resources unavailable to the average researcher. On the
other hand, one clear advantage of volunteer computing frameworks is
its availability: in principle, any one can request volunteers. How
many she is able to gather, on the other hand, is an issue that we
will try to approach in this paper.

Besides, since Amazon started selling EC2 several years ago, reliable
and scalable computing resources are also available for a low price
and on demand. Recently, Google has also refurbished its offering
lowering their prices. This means that the conjunction of free or
low-cost cloud computing engines, volunteer computing systems and
untapped capability of other people's desktop systems can be used for creating
massive, or at least potentially massive, distributed computing
experiments. These experiments can be easily set up hosting the
sources and data in open-source-friendly
repository sites like GitHub\footnote{\url{http://www.github.com}} and deployed automatically to Platform as
a Service (PaaS) hosts such as
Heroku\footnote{\url{https://www.heroku.com/}} or
OpenShift\footnote{\url{https://www.openshift.com/}} or even
Infrastructure as a Service (IaaS) sites such as Microsoft's Azure or
Amazon's EC2.

@@

These are best practices that have been followed in the experiments in
this paper, that first presents a new version of a platform for
distributed evolutionary algorithms (EAs) using the browser and a free (as
in free beer) backend host, and second, shows the result from a
statistical point of view in order to put the basis for a model of the
meta-computer obtained by joining all volunteers and the free backend
used for the experiment. This has been one of the criterion when
choosing the different cloud services (hosting and logging, mainly):
the fact that they are free, and sustainably so
\cite{Merelo2014,jj:idc:lowcost}.

@@

\section{Results}
\label{sec:res}
%
\begin{figure}[h!tb]
        \centering
        \includegraphics[width=0.95\linewidth]{img/times-base.png}
        \caption{Histogram of time needed to find the solution in the
          command-line client case, with (Y) or without restart (N)
          when the solution is found.}\label{fig:base}
\end{figure}
%
In order to establish a baseline, we wrote an evolutionary algorithm
in node using the NodEO framework \cite{DBLP:conf/gecco/GuervosVGES14}
in the command line. Conditions, and, indeed, the code, were the same except for the
population, which was set to 512 individuals, and the termination
condition, which included finishing if a sufficient amount of
generations had passed without finding the solution. We did 50 experiments restarting the server,
which contains the pool of best individuals every 100 generations,
when the client had finished without finding the solution, or not;
this second case is more similar to the asynchronous nature of the
browser-based experiment. In fact, no difference was found (using
Wilcoxon test) and the average time to solution was 34.21 seconds, the
distribution of times does not follow a normal, however, and is
plotted in Figure \ref{fig:base}.

@@

At the time of this writing, 23 runs were done in OpenShift and one in  
% para trabajos futuros, estas cifras de cuántas ejecuciones se han completado 
% quizás habría que presentarlas de forma diferente. Quizás indicando el tiempo
% total de "puesta online" del sistema, cómo se publicitó para atraer gente,
% y entonces ahí contar las estadísticas de cuánta gente partició, cuántas
% ejecuciones se terminaron, etc.  [Pedro]
% se terminaron todas. Lo de cuanta gente por ejecución (cuantas IPs,
% en realidad) y cuanto tardaron no lo he llegado a analizar, aunque
% tengo los datos. Se deja para trabajo futuro, puedes añadirlo a un
% issue - JJ
Heroku. The Openshift URL was mentioned more frequently than the other
so that accounts for the popularity, not any particular problem with
the algorithm in one or the other. That is why the data that we have
obtained and used below, and which is, as usual, available with a free
license, is mainly processed from the log files stored in Loggly and
in OpenShift. These 23 experiments took, all together, a number of hours, so
that in many cases the nodes were not accessing at the same time, but
sequentially, to the server. The average time for these experiments
was 466.4 seconds, with the median on 374 seconds, that is, around 20
times as much as the baseline experiment mentioned above. It should be
noted, however, that our intention was not to use distributed
evolutionary computing to optimize the running time, but the resources
devoted to them. This measure indicates, at any rate, that there is
some room for improvement and future research should go, among others,
in that direction.

This means that the island model is, in
fact, behaving partly as a pool-based evolutionary algorithm
\cite{sofea:evopar2012} with some results, the best from each
generation, available to future runs. This is clearly noticed when
visualizing the algorithm: its behavior changes when it gets its first
chromosome from the pool, in many cases improving a lot.
